---
title: " SAGB Article 1: What is peak age in professional tennis?"
author: "Natalie Brewer"
date: "2023-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load relevant libraries
library(ggplot2)
library(dplyr)
library(readr)
library(MASS)
```

## Import data

Import data from Jeff Sackman tennis database, Ultimate Tennis Statistics, and ESPN.
```{r}
# Player names and basic information, including birth date
player_data <- read.table("/Users/nataliebrewer/Desktop/SAGB/Article 1/atp_players.csv", sep=",")
player_df <- data.frame(player_data)
# Remove extra data and rename
player_df <- player_df[-1,] #Remove row of labels
colnames(player_df) <- c("PlayerID", "First_Name", "Last_Name", "Hand", "Birthdate", "Country", "Height", "Wiki_ID")  #Rename columns
player_df$Full_Name <- paste(player_df$First_Name, player_df$Last_Name, sep=" ")  #Combine first and last name columns into one
player_df$Birthdate <- as.Date(player_df$Birthdate, format="%Y%m%d")  #Convert birthdates to date format
head(player_df)



# Peak ELO and ATP rankings and age when achieved
peakELO_data <- read.table("/Users/nataliebrewer/Desktop/SAGB/Article 1/PeakEloRatings.csv", sep=",")
peakELO_df <- data.frame(peakELO_data)
# Remove extra data and rename 
peakELO_df <- peakELO_df[-1,-c(1,3,4,5,10,11,12)] #Remove unneeded columns and label row
colnames(peakELO_df) <- c("Name", "Peak_ELO", "Best_Rank", "Best_Rank_Date", "Peak_ELO_Date", "Peak_ELO_Age") #Rename columns
# Convert data types
peakELO_df$Peak_ELO_Age <- as.numeric(peakELO_df$Peak_ELO_Age) #Convert ages to numbers
peakELO_df$Peak_ELO_Date <- as.Date(peakELO_df$Peak_ELO_Date) #Convert peak Elo dates to dates
peakELO_df$Best_Rank_Date <- as.Date(peakELO_df$Best_Rank_Date) #Convert peak ATP dates to dates
peakELO_df$Best_Rank_Age <- as.integer(difftime(peakELO_df$Best_Rank_Date, peakELO_df$Peak_ELO_Date, units="weeks")  / 52.25 + peakELO_df$Peak_ELO_Age)   #Convert ages to numbers
head(peakELO_df)


# Grand Slam winners for 4 events each year 
GS_data <- read.table("/Users/nataliebrewer/Desktop/SAGB/Article 1/GrandSlamHistory.csv", sep=",")
GS_df <- data.frame(GS_data)
# Rename columns, remove extra data, and reorder by ascending year
colnames(GS_df) = c("Year", "Tournament", "Winner", "Runner_Up") #Rename columns
GS_df <- GS_df[-1,] #Remove label row
GS_df <- GS_df[order(GS_df$Year),] #Reorder by year
GS_df <- GS_df[GS_df$Year >1950,] #Remove years prior to 1950
# Remove rows with other tournaments
GS_df <- GS_df[GS_df$Tournament %in% c("U.S. Open", "Wimbledon", "French Open","Australian Open"),]
#Add column with tournament date (estimated, but correct up to 2 weeks)
GS_df <- GS_df %>% 
  mutate(Tourn_Date = case_when(
    Tournament == "U.S. Open" ~ as.Date(paste(GS_df$Year,"09-09", sep="-")),
    Tournament == "Wimbledon" ~ as.Date(paste(GS_df$Year,"07-01", sep="-")),
    Tournament == "French Open" ~ as.Date(paste(GS_df$Year,"06-09", sep="-")),
    Tournament == "Australian Open" ~ as.Date(paste(GS_df$Year,"02-01", sep="-")),
    TRUE ~ NA
  ))
GS_df$Year <- as.numeric(GS_df$Year) #Convert year to number
head(GS_df)

```

## Peak ATP and Elo Ratings

Generate peak ATP ranking and peak Elo rating charts.
```{r}
# Generate plots
ggplot(peakELO_df, aes(x=Peak_ELO_Age)) +
  geom_histogram(aes(y=after_stat(density)), binwidth=1, width=0.95, fill="#c3f734", color="white", alpha=0.7) +
  geom_density(aes(y=after_stat(density)), color="blue", linewidth=1.5) +  
  labs(title="Age when reached peak Elo rating", x="Peak Elo Age", y="Density") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )

ggplot(peakELO_df, aes(x=Best_Rank_Age)) +
  geom_histogram(aes(y=after_stat(density)), binwidth=1, fill="#c3f734", color="white", alpha=0.7) +
  geom_density(aes(y=after_stat(density)), color="blue", linewidth=1.5) +  
  labs(title="Age when reached peak ATP ranking", x="Peak ATP Ranking Age", y="Density") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )

```



Next, generate peak Elo rating chart, separated by decade.

For the sake of simplicity, I've combined the 2010s with the first 3 years of the 2020s.
```{r}
# Add a column for decade.
peakELO_df$Decade <- NA
# Loop through the rows and categorize by decade
for (i in 1:nrow(peakELO_df)) {
  date <- peakELO_df$Peak_ELO_Date[i]
  if (as.Date("1980-01-01") <= date & date < as.Date("1990-01-01")) {
    peakELO_df$Decade[i] <- "1980s"
  } else if (as.Date("1990-01-01") <= date & date < as.Date("2000-01-01")) {
    peakELO_df$Decade[i] <- "1990s"
  } else if (as.Date("2000-01-01") <= date & date < as.Date("2010-01-01")) {
    peakELO_df$Decade[i] <- "2000s"
  } else if (as.Date("2010-01-01") <= date & date < as.Date("2024-01-01")) {
    peakELO_df$Decade[i] <- "2010-23"
  }
}

# Generate plot for decade-separated peak Elo densities
ggplot(peakELO_df %>% filter(Peak_ELO_Date >= as.Date("1980-01-01")), aes(x = Peak_ELO_Age, y = after_stat(density), color = Decade))  +
  geom_density(linewidth = 1.5) +
  labs(title = "Age when reached peak Elo rating", x = "Peak Elo Age", y = "Density") +
  theme_minimal() +
  scale_color_manual(
    values = c("1980s" = "#f7b334", "1990s" = "#f7f134", "2000s" = "#c3f734", "2010-23" = "#34f758"), 
    name = "Decade"  # Title for the legend
  )+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )
```


## Grand Slam Winners Analysis

```{r}
# Combine player data and grand slam data frames, align winners to their player profile
GS_df <- GS_df %>%
  left_join(player_df, by = c("Winner" = "Full_Name")) %>%
  mutate(Winner_Age = as.numeric(difftime(Tourn_Date, Birthdate, units = "weeks") / 52.25))

# Remove rows with no player age data and duplicates
GS_df <- GS_df[complete.cases(GS_df$Winner_Age),]
GS_df <- GS_df[-c(90, 91), ]
GS_df <- GS_df[-82, ]
head(GS_df)

# Separate data frames for each tournament
usopen_df <- GS_df[GS_df$Tournament == "U.S. Open", ]
wimb_df <- GS_df[GS_df$Tournament == "Wimbledon", ]
french_df <- GS_df[GS_df$Tournament == "French Open", ]
aus_df <- GS_df[GS_df$Tournament == "Australian Open", ]

# Average age of winner each year
avg_df <- data.frame(Year = 1951:2023)
avg_age_vec <- c()
count <- 0
total <- 0

# Create data frame with averaged age of winners for each year
for (i in avg_df$Year) {
  if (i %in% usopen_df$Year) {
    count <- count + 1
    total <- total + usopen_df[usopen_df$Year == i, "Winner_Age"]
  }
  if (i %in% wimb_df$Year) {
    count <- count + 1
    total <- total + wimb_df[wimb_df$Year == i, "Winner_Age"]
  }
  if (i %in% french_df$Year) {
    count <- count + 1
    total <- total + french_df[french_df$Year == i, "Winner_Age"]
  }
  if (i %in% aus_df$Year) {
    count <- count + 1
    total <- total + aus_df[aus_df$Year == i, "Winner_Age"]
  }
  avg_age_vec <- c(avg_age_vec, total/count)
  count <- 0
  total <- 0
}

avg_df$Age <- avg_age_vec
```

First, we look at the ages of the winners over time.

```{r}
# Line plot over time, each color is a different event
ggplot() +
  geom_point(data=usopen_df,aes(x=Year, y=Winner_Age), color="red") +
  geom_line(data=usopen_df,aes(x=Year, y=Winner_Age, group=1), color="red") +
  geom_point(data=wimb_df,aes(x=Year, y=Winner_Age), color="blue") +
  geom_line(data=wimb_df,aes(x=Year, y=Winner_Age, group=1), color="blue") +
  geom_point(data=french_df,aes(x=Year, y=Winner_Age), color="green") +
  geom_line(data=french_df,aes(x=Year, y=Winner_Age, group=1), color="green") +
  geom_point(data=aus_df,aes(x=Year, y=Winner_Age), color="yellow") +
  geom_line(data=aus_df,aes(x=Year, y=Winner_Age, group=1), color="yellow") +
  labs(title="Winners of Grand Slams over time", x="Year",y="Age of Champion") +
  scale_y_continuous(limits =c(0,40), breaks=seq(0,45,by=5)) +
  scale_x_continuous(limits =c(1950,2025), breaks=seq(1950,2025,by=5))


# Line plot over time, all events averaged
ggplot() +
  geom_smooth(data = avg_df, aes(x = Year, y = Age), method = "lm", se = FALSE, color = "#c3f734") +  #Regression line
  geom_point(data=avg_df,aes(x=Year, y=Age), color="blue") +
  geom_line(data=avg_df,aes(x=Year, y=Age, group=1), color="blue") +
  labs(title="Average age of Grand Slam winners each year", x="Year",y="Average Age of Champions") +
  scale_y_continuous(limits =c(15,35), breaks=seq(0,45,by=5)) +
  scale_x_continuous(limits =c(1950,2023), breaks=seq(1950,2023,by=10))+
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )
```

Next we look at all the winners of the events to see which ages are most common among them.

```{r}
# All Grand Slams together histogram
ggplot(GS_df, aes(x=Winner_Age)) +
  geom_histogram(aes(y=after_stat(density)), binwidth = 1, fill="#f7b334", color="white") +
  #stat_function(fun=dgamma, args=list(shape=fit$estimate["shape"], rate=fit$estimate["rate"]), color="blue", linewidth=1.5) +
  labs(title="Grand Slam Winner Ages", x="Age", y="Density") +
  scale_x_continuous(breaks=seq(0,45, by=1)) + 
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )

# US Open histogram
ggplot(usopen_df, aes(x = Winner_Age)) +
  geom_histogram(binwidth = 1, fill = "#c3f734", color="white") +
  labs(title = "US Open Winner Ages", x = "Age", y = "Number of Winners") +
  scale_x_continuous(breaks = seq(0, 45, by = 2)) +
  scale_y_continuous(breaks = seq(0, 15, by = 5)) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )

# Wimbledon histogram
ggplot(wimb_df, aes(x=Winner_Age)) +
  geom_histogram(binwidth = 1, fill ="#c3f734", color="white") +
  labs(title="Wimbledon Winner Ages", x="Age", y="Number of Winners") +
  scale_x_continuous(breaks=seq(0,45, by=2)) +
  scale_y_continuous(breaks = seq(0, 15, by = 5)) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )

# French Open histogram
ggplot(french_df, aes(x=Winner_Age)) +
  geom_histogram(binwidth = 1, fill ="#c3f734", color="white") +
  labs(title="French Open Winner Ages", x="Age", y="Number of Winners") +
  scale_x_continuous(breaks=seq(0,45, by=2)) +
  scale_y_continuous(breaks = seq(0, 15, by = 5)) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )

# Australian Open histogram
ggplot(aus_df, aes(x=Winner_Age)) +
  geom_histogram(binwidth = 1, fill ="#c3f734", color="white") +
  labs(title="Australian Open Winner Ages", x="Age", y="Number of Winners") +
  scale_x_continuous(breaks=seq(0,45, by=2)) + 
  scale_y_continuous(breaks = seq(0, 15, by = 5)) +
  theme_minimal()+
  theme(
    axis.text.x = element_text(margin = margin(t = 5, b = 15)),
    axis.text.y = element_text(margin = margin(r = 5, l = 15), hjust = .5)
  )
```
